 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-07-29 paper.Rmd
update paper and include videos

There has been a rapid uptake in the use of non-linear dimensionality reduction
(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
the natural sciences as part of cluster orientation and dimension reduction
workflows. The appropriate use of these methods is made difficult by their
complex parameterisations and the multitude of decisions required to balance
the preservation of local and global structure in the resulting visualisation.
We present a visual analytics framework for the pragmatic usage of NLDR methods
by combining them with a technique called the tour. A tour is a sequence of
interpolated linear projections of multivariate data onto a lower dimensional
space. The sequence is displayed as a dynamic visualisation, allowing a user to
see the shadows the high-dimensional data casts in a lower dimensional view. By
linking the tour to a view obtained from an NLDR method, we can preserve global
structure and through user interactions like spatial linked brushing observe
where the NLDR view may be misleading. We display several case studies from
single cell transciptomics and image processing, that shows our approach is
useful for cluster orientation tasks.  The implementation of our framework is
available as an R package called **liminal** available at
https://github.com/sa-lee/liminal.

@@

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of both data cleaning /
pre-processing and visualisation. Methods to embed data from a high-dimensional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and de-noise data
for downstream analysis (thereby nullifying the 'curse of dimensionality').

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow.  @Brehmer2014-hk characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters. Both of these
workflow steps rely on the embedding being representative the original high
dimensional dataset, and becomes much more difficult when there is no underlying
ground truth.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they also noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complementary approach for visualizing structure in high dimensional data is 
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as a
animated visualization.Given the dynamic nature of the tour, user interaction is 
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
exploring statistical model fits and by @Buja1996-fk for exploring the space of 
factorial experimental designs.

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to an estimated
pairwise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
The idea behind this is that non-linear embedding methods should preserve 
locality and that points that are 'close' to each other on the embedded view
should be close according to the estimated pairwise distance. 
Similarly, the TensorFlow embedding projector is a web interface to 
running some non-liner embedding methods live in the browser and 
provides interactions  to colour points, and select nearest neighbours
[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
guided and modified form of the  t-SNE algorithm, that allows users to modified 
optimisation parameters in real-time.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called **liminal** (Available: https://github.com/sa-lee/liminal).

## Dimensionality Reduction Background

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. Here we restrict our attention to two recent methods that
are commonly used in the literature: t-distributed stochastic neighbour embedding 
(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
mention several other techniques.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pairwise distances (inner products) in $Y$ 
approximate the pairwise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pairwise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such asdiffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pairwise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

The t-SNE algorithm estimates
the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pairwise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

@@

Next, the target low-dimensional space, $Y$, pairwise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pairwise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pairwise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

@@

Another recent method UMAP has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
Its main differences are that it uses a different cost function (cross entropy)
which is optimized using stochastic gradient descent and defines a different
kernel for similarities in the low dimensional space. Due to it's computational
speed it's possible to generate UMAP embeddings in more than three dimensions.
It appears to suffer from the same perceptual issues as t-SNE, however it
supposedly preserves global structure better than t-SNE [@Coenen2019-to].

@@

We propose using tours as part of an analyst's workflow when using embedding
methods like t-SNE. We have made extensive use of ensemble graphics, that is
aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
As we will see in the case studies, this allows analysts to quickly compare 
views from embedding methods and allows them to see how the embedding method 
alters the global structure of their data. Using ensembles allows the use of 
interaction techniques, that allow analysts to perform cluster orientation 
tasks via linking multiple views. This approach allows our interface, to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an initial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In our interface, brushing will pause the tour path, allowing
users to identify 'interesting' projections. The domain of the axis scales from
running a tour is called the half range, and is computed by rescaling the input
data onto hyper-dimensional unit cube. We bind the half range to a mouse wheel
event, allowing a user to pan and zoom on the tour view dynamically. This is
useful for peeling back dense clumps of points to reveal structure.

@@

We have combined the tour view in a side by side layout with a scatter plot
view that represents the output of an embedding algorithm.  These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labelled according to some discrete structure, i.e. cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

@@

We propose this can be
done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
of brushing acts directly on $k$-nearest neighbours ($l$-nn) graph computed from a
reference dataset, when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view. In place of highlighting, one could use opacity
or binned colour scales to encode distances or ranks instead of the 
neighbouring points, or by highlighting. Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views. This is left for 
future implementations of **liminal**.

@@

We have implemented the above design as an open source R package called
**liminal** [@baseR]. The package allows analysts to construct concatenated
visualisations, via the Vega-Lite grammar of interactive graphics (using
`vegawidget` package) and provides an interface for constructing brushes and
manipulating tour paths using the `shiny` and `tourr` packages. **liminal** also
provides a stand-alone interface to the tour, in addition to the linked
scatterplot interfaces discussed above. Below we display the **liminal** API, and
our approach to generating tour paths and user interactions.

Here we provide an overview of the mathematical aspects of the tour for
readers who are unfamiliar with the technique before discussing the 
computational aspects of how **liminal** interfaces with them. The tour
is emphatically not a dimension reduction technique but rather a visualisation 
technique that is grounded in mathematical theory, that is able to ascertain 
the shape and global structure of a dataset via inspection of the subspace 
generated by the set of low-dimensional projections. In fact, in some
of the case studies below, we  first sphere our data via principal 
components to reduce dimensionality of our input.

Like, when using DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence 'interesting' 
projections as quantified by an index function. While **liminal** is able to 
visualise guided tours, our focus in the case studies uses the grand tour to
see global structure in the data.

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatterplot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flipbook" 
animation. Like with a flipbook, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in **liminal**
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

@@

The embedding and tour views are linked together via rectangular brushes;
when a brush is active, points will be highlighted in the adjacent view. 
Because the tour is dynamic, brush events that become active will pause the
animation, so that a user can interrogate the current view. By default,
brushing on the embedding view will produce a one-to-one linking with the tour
view (the special case of $k=1$ nearest neighbours brush). In future
implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.  If desired, the distance metric used to estimate nearest neighbours can by chosen by the user. 
The user may provide a reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
independently of the tour or embedding views. The estimation of neighbours
is done with the `FNN` package for datasets with less than 4096 observations
and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
with the interface, the brush regions are returned for downstream analysis.

@@

The next section steps through case studies cases of our approach using simulated
and real data.  The videos presented show the visual 
appearance of the **liminal** interface, and how we can interact with the tour
via the controls previously described.
In the first case study, we look at three different examples
where t-SNE is applied to three simulations where cluster structure and geometry
of the underlying data is known.

In the second case study, we apply our approach to clustering 
the mouse retina data from @Macosko2015-ot, and use apply the tour
to the process of verifying marker genes.

<!--
Finally, in the last case study we
explore a subset of the QuickDraw dataset, we use **liminal** to identify
instances of images from different labels that appear similar.
-->

### Case Study 1: Exploring simulation data with tours and t-SNE

To begin we look at simulated datasets that reproduce known facts
about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
covariance matrix. We then computed a t-SNE layout with default settings 
using the **Rtsne** package [@Rtsne],
and set up the **liminal** linked interface with grand tour on the 10-$d$
observations (video - ).

<embedded video liminal gaussian>

We see in the video that t-SNE has correctly split out each cluster and
laid them out in a star like formation. This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout. This can be seen from our brushing action on the t-SNE side.

<embedded video liminal hierachcial >

We repeat the process of simulating Gaussian clusters in video <>, but this time
we have a cluster that has more granular structure. This time we have two
Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
with hierachcial structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
Again t-SNE has correctly identified the sub-clusters but their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view. We see in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent. When we brush the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

>>>>

```{r setup-videos, cache = FALSE}
vid_lookup <- tibble::tribble(
  ~key, ~url, ~path,
  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("img", "liminal-screenshot-gaussian.png"),
  "hierarchical", "https://player.vimeo.com/video/439635905", here::here("img", "liminal-screenshot-hierarchical.png"),
  "trees-01", "https://player.vimeo.com/video/439635892", here::here("img", "liminal-screenshot-trees-01.png"),
  "trees-02", "https://player.vimeo.com/video/439635863", here::here("img", "liminal-screenshot-trees-02.png"),
  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("img", "liminal-screenshot-mouse-01.png"),
  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("img", "liminal-screenshot-mouse-02.png")
)

display_vid_html <- function(key, lookup) {
  video <- lookup[lookup$key == key, ]
  knitr::include_url(video[["url"]])
}

display_screenshot <- function(key, lookup) {
  img <- lookup[lookup$key == key, ]
  knitr::include_graphics(img[["path"]])
}

```

# Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings {#ch:tsne}

There has been a rapid uptake in the use of non-linear dimensionality reduction
(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
the natural sciences as part of cluster orientation and dimension reduction
workflows. The appropriate use of these methods is made difficult by their
complex parameterisations and the multitude of decisions required to balance
the preservation of local and global structure in the resulting visualisation.
We present visual diagnostics for the pragmatic usage of NLDR methods
by combining them with a technique called the tour. A tour is a sequence of
interpolated linear projections of multivariate data onto a lower dimensional
space. The sequence is displayed as a dynamic visualisation, allowing a user to
see the shadows the high-dimensional data casts in a lower dimensional view. By
linking the tour to a view obtained from an NLDR method, we can preserve global
structure and through user interactions like linked brushing observe
where the NLDR view may be misleading. We display several case studies from
both simulated and real data from single cell transcriptomics, that shows our
approach is useful for cluster orientation tasks.  The implementation of our 
framework is available as an R package called **liminal** available at
https://github.com/sa-lee/liminal.

@@

High dimensional data is increasingly prevalent in the natural sciences and
beyond but presents a challenge to the analyst in terms of data cleaning,
pre-processing and visualisation. Methods to embed data from a high-dimensional
space into a low-dimensional one now form a core step of the data analysis
workflow where they are used to ascertain hidden structure and de-noise data
for downstream analysis .

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow. @Brehmer2014-hk characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters. Both of these
workflow steps rely on the embedding being representative of the original high
dimensional dataset, and becomes much more difficult when there is no underlying
ground truth.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they also noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2D scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complementary approach for visualizing structure in high dimensional data is 
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as a
animated visualization [@Asimov1985-vp; @Buja1986-zr]. Given the dynamic nature
of the tour, user interaction is  important for controlling and exploring the 
visualisation: the tour has been used previously by @Wickham2015-cx and  
exploring statistical model fits and by @Buja1996-fk for exploring the space of 
factorial experimental designs.

While there has been much work on the algorithmic details of
embedding methods, there are relatively few tools designed to assist users
to interact with these techniques: when is an embedding sufficient for the
task at hand? Several
interactive interfaces have been proposed for evaluating or using embedding
techniques. @Buja2008-fn used tours to guide analysts during the optimisation
of multidimensional scaling methods by extending their interactive visualisation
software called **XGobi** and **GGobi** into a new tool called **GGvis** 
[@Swayne1998-uq; @Swayne2003-qd; @Swayne2004-ua].
Their interface allows the analyst to dynamically modify and check whether
an MDS configuration has preserved the locality and closeness
of points between the configuration and the original data.
@Ovchinnikova2020-sy created the **Sleepwalk**  interface for checking
non-linear embeddings in single cell RNA-seq data. It provides a click and 
highlight visualisation for colouring points in an embedding according to an 
estimated pairwise distance in the original high-dimensional space.
Similarly, the **TensorFlow**  embedding projector is a web interface to 
running some non-liner embedding methods live in the browser and 
provides interactions to colour points, and select nearest neighbours
[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
guided and modified form of the  t-SNE algorithm, that allows users to modify 
optimisation parameters in real-time.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. Instead, we implement a more pragmatic workflow by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks.

The rest of the paper is organised as follows. The next section provides
background on dimension reduction methods, including an overview of the tour.
Then we describe the visual design of **liminal**, followed by implementation
details. Next we provide case studies that show how our interface assists
in using embedding methods. Finally, we describe the insights gained by using
**liminal** and plans for extensions to the software.

## Overview of Dimension Reduction

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. Here we restrict our attention to two recent methods that
are commonly used in the literature: t-distributed stochastic neighbour embedding 
(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
mention several other techniques [@Maaten2008-sk; @McInnes2018-co].

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

The t-SNE algorithm estimates
the pair-wise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom [@Maaten2008-sk]. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

@@

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

@@

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pair-wise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pair-wise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

@@

Another recent method, UMAP, has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
Its main differences are that it uses a different cost function (cross entropy)
which is optimized using stochastic gradient descent and defines a different
kernel for similarities in the low dimensional space. Due to its computational
speed it is possible to generate UMAP embeddings in more than three dimensions.
It appears to suffer from the same perceptual issues as t-SNE, however it supposedly preserves global structure better than t-SNE [@Coenen2019-to].

The tour is a visualisation technique that is grounded in 
mathematical theory, and is able to ascertain 
the shape and global structure of a dataset via inspection of the subspace generated by the set of low-dimensional projections [@Asimov1985-vp; @Buja1986-zr].

Like, when using other DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections.  In practice, we first sphere 
our data via principal components to reduce dimensionality of $X$ prior to 
running the tour. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence of 'interesting' 
projections as quantified by an index function [@Cook1995-bi]. While our software, 
**liminal** is able to  visualise guided tours, our focus in the case studies 
uses the grand tour to see global structure in the data.

@@

Tours provide a supportive visualisation to NLDR graphics, and 
can be easily incorporated into an analysts workflow with our software package, 
**liminal**.  Our interface allows analysts to quickly compare 
views from embedding methods and see how an embedding method 
preserves or alters the geometry of their data. Using multiple concatenated and
linked views with the tour enhances interaction techniques, and allows 
analysts to perform  cluster orientation tasks via linked highlighting and 
brushing [@McDonald1982-ew; @Becker1987-gd]. 
This approach allows our interface to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

To investigate latent structure and the shape of a high dimensional dataset, a
tour can be run without the use of an external embedding. It is often useful to
first run principal components on the input as an initial dimension reduction
step, and then tour a subset of those components instead, i.e. by selecting
them from a scree plot. The default tour layout is a scatter plot with an axis
layout displaying the magnitude and direction of each basis vector. Since the
tour is dynamic, it is often useful to be able to pause and highlight a
particular view. In addition to pause, play and reset buttons, brushing will 
pause the tour path, allowing users to identify 'interesting' projections. The
domain of  the axis scales from running a tour is called the half range, and is 
computed by rescaling the input data onto hyper-dimensional unit cube. We bind 
the half range to a mouse wheel event, allowing a user to pan and zoom on the 
tour view dynamically. This is useful for peeling back dense clumps of points 
to reveal structure.

@@

We have combined the tour view in a side by side layout with a scatter plot
view as has been done in previous tour interfaces **XGobi** and **DataViewer** 
[@Buja1986-ku; @Swayne1998-uq]. These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labelled according to some discrete structure, i.e. cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

@@

These checks are done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. Similarly, diffusions
are identified from linked brushing on the tour view, highlighting points
in the embedding view.

@@

We have implemented the above design as an open source R package called
**liminal** [@r-liminal]. The package allows analysts to construct concatenated
visualisations, drawn with the **Vega-Lite** grammar of interactive graphics via
the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides 
an interface for constructing linked and stand alone interfaces for 
manipulating tour paths via  the **shiny** and **tourr** packages 
[@Chang2020-bq; @Wickham2011-st].

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatter plot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flip book" 
animation. Like with a flip book, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in **liminal**
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

@@

The embedding and tour views are linked together via rectangular brushes;
when a brush is active, points will be highlighted in the adjacent view. 
Because the tour is dynamic, brush events that become active will pause the
animation, so that a user can interrogate the current view. By default,
brushing on the embedding view will produce a one-to-one linking with the tour
view. For interpreting specific combinations of clusters, the multiple
guides on the legend can be selected in order to see their relative orientations.
The interface is constructed as a **shiny** gadget specifically designed for
interactive data analysis. Selections such as brushing regions and the current
tour path are returned after the user clicks done on the interface and become
available for further investigation.

@@

The next section steps through case studies cases of our approach using simulated
and real data.

The first three case studies use simulations where the cluster structure and
geometry of the underlying data is known. We start with a simple example
where we generated spherical clusters that are embedded well by t-SNE. Then
we move onto more complex examples where the tour provides insight, such
as clusters that have substructure and where there is more complex geometry
in the data.

In the final case study, we apply our approach to clustering 
the mouse retina data from @Macosko2015-ot, and apply the tour
to the process of verifying marker genes that separate clusters.

The videos presented show the visual 
appearance of the **liminal** interface, and how we can interact with the tour
via the controls previously described. Both code and videos available for 
reproducing this paper are available at https://github.com/sa-lee/paper-liminal.

### Case Study 1: Exploring spherical Gaussian clusters

To begin we look at simulated datasets that reproduce known facts
about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
covariance matrix. We then computed a t-SNE layout with default settings 
using the **Rtsne** package [@Rtsne],
and set up the **liminal** linked interface with grand tour on the 10-$d$
observations.

(ref:liminal-gaussian) Screenshots of the **liminal** interface applied to well clustered data, a video of the tour animation is available at https://player.vimeo.com/video/439635921. *(a)*: t-SNE (left) and tour (right) in side by side layout coloured by cluster. *(b)*: brushing the red cluster on the t-SNE view, shows the corresponding points on the tour.

```{r gaussian, fig.cap = "(ref:liminal-gaussian)"}
display_screenshot("gaussian", vid_lookup)
```

```{r gaussian-h, eval = knitr::is_html_output()}
display_vid_html("gaussian", vid_lookup)
```

We see in figure \@ref(fig:gaussian) and linked video that t-SNE has 
correctly split out each cluster and laid them out in a star like formation. 
This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout (figure \@ref(fig:gaussian)a). 
This can be confirmed via brushing the t-SNE view (figure \@ref(fig:gaussian)b).

### Case Study 2: Exploring spherical Gaussian clusters with hierarchical structure

We repeat the process of viewing Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in $10-d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
Again t-SNE has correctly identified the sub-clusters but their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

(ref:liminal-hierarchical) Screenshots of the **liminal** interface applied to sub-clustered data, a video of the tour animation is available at https://player.vimeo.com/video/43963590. *(a)* t-SNE has identified and spread out the clusters with substructure. *(b)* Using linked highlighting and brushing, the tour view shows that sub-clusters are much closer together in high dimensional space than the t-SNE view implies.

```{r hierarchical, fig.cap="(ref:liminal-hierarchical)"}
display_screenshot("hierarchical", vid_lookup)
```

```{r hierarchicall-h, eval = knitr::is_html_output()}
display_vid_html("hierarchical", vid_lookup)
```

### Case Study 3: Exploring data with piecewise linear structure
